---
title: "lecture 7"
author: "Anuhya Balineni"
output: html_document
date: "2025-02-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(nnet)
library(ISLR2)
```

# 1. INTRODUCTION

\< Introduce the models being used\>

# 2. Data

\< Describe the data\>

```{r}
data = Default
str(data)
```

## 2.1 Visualizing the data

### 2.1.1 Distribution of Balance

\<what does this figure mean?\>

```{r balance distribution}
ggplot(data, aes(x =balance, fill=default)) +
  geom_histogram(bins = 30, alpha = 0.7, position = 'identity' ) +
  labs(title = "Distribution of balance by Default Status",
       x = "Balance",
       y = "Count?")
 
```

### 2.1.2 Distribution of Income

\<what does this figure mean?\>

```{r}
ggplot(data, aes(x = income, fill = default)) +
  geom_histogram(bins = 30, alpha = 0.7, position = 'identity') +
  labs(title = "distribution of Income by Default Status",
       x = "Income",
       y = "Count")
```

```{r}
ggplot(data, aes(x = income, fill = student)) +
  geom_histogram(bins = 30, alpha = 0.7, position = 'identity') +
  labs(title = "distribution of Income by Default Status",
       x = "Income",
       y = "Count")
```

### 2.1.3 Student Status by Default

```{r}
ggplot(data, aes(x = student, fill = default)) +
  geom_bar(position ='dodge') +
  labs(title = "distribution of Income by Student Status",
       x = "Student",
       y = "Count")
```

# 4. Logistic Regression

### 4.1 Fitting the Model

\<Describe Logistic Regression\>

```{r}
logit_model = glm(default ~ balance, data = data, family = binomial)
summary(logit_model)
```

```{r}
data$predicted_prob = predict(logit_model, type = "response")
head(data)
```

### 4.2 Evaluate Model Performance

\<talk about our model and evaluation metrics\>

```{r}
threshold = 0.5
data$predicted_default = ifelse(data$predicted_prob > threshold, "yes", "No")
conf_matrix = table(data$predicted_default, data$default)
conf_matrix
```

```{r}
accuracy = sum(diag(conf_matrix)) / sum(conf_matrix)
accuracy
```

# 5. Multiple Logistic Regression

## 5.1 Fitting the Model

Here, we will include an **interaction term** between 'income' and 'student' that allows the effect of 'income'

```{r}
logit_mult_model = glm(default ~ balance + income * student, data=data, 
family=binomial)
summary(logit_mult_model)
```

## 5.2 Evaluating the Model

\<talk about evaluation metrics / interpretation\>

```{r}
data$mult_predicted_prob = predict(logit_mult_model, type = "response")
data$mult_predicted_default = ifelse(data$mult_predicted_prob > threshold, "Yes", "No")
conf_matrix_mult = table(data$mult_predicted_default, data$default)
conf_matrix_mult
```

```{r}
accuracy_mult = sum(diag(conf_matrix_mult)) / sum(conf_matrix_mult)
accuracy_mult
```

# 6. Multinomial Logistic Regression

## 6.1 Load the Data

```{r}
data2 = Carseats
data2$SalesCategory = cut(data2$Sales, breaks = 3, labels = c("Low", "Medium", "High"))
```

```{r}
multi_model = multinom(SalesCategory ~ Price + Income +Advertising,
data = data2)
summary(multi_model)
```

## 6.2 Make Predictions

```{r}
data2$nomial_predicted_salesCat = predict(multi_model)
head(data2)
```

## 6.3 Evaluate Model

```{r}
conf_matrix_multi = table(data2$nomial_predicted_salesCat, data2$SalesCategory)
conf_matrix_multi
```

```{r}
accuracy_multi = sum(diag(conf_matrix_multi)) / sum(conf_matrix_multi) 
accuracy_multi
```

# Assignment Section

### Background

Diabetes is a chronic disease affecting millions of individuals worldwide. Early detection through predictive modelling can help guide prevention and treatment. in this assignment , you will use logistic regression to predict whether an individual has diabetes using basic health information.

We will use the Pima Indians Diabetes Dataset, a commonly used dataset in health informatics available from the UCI Machine Learning Repository and built into the mlbench R package.

## Simple Logistic Regression

```{r}
#install.packages("mlbench")
library(mlbench)
data("PimaIndiansDuabetes")
df = PimaIndiansDiabetes
```

### Data Exploration and Summary Figures

### Fit a Simple Logistic Regression Model (Train & Test Split)

\< Fit a logistic regression using glucose as a predictors of diabetes \>

### Interpret Coefficients & Apply the Model for Prediction on the test Data

## Multiple Logistic Regression
